---
title: Morphometric distribution of depth and basin shape in Aotearoa New Zealand lakes
format: 
  revealjs:
    theme: dark
    logo: www/limnotrack_border.jpg
    footer: "Footer text"
    template-partials:
      - title-slide.html
author:
  - name: Tadhg Moore
    affiliation: Limnotrack
    orcid: 0000-0002-3834-8868
    email: tadhg@limnotrack.com
  - name: Chris McBride
    affiliation: Limnotrack
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup

library(sf)
fenz_depths <- readRDS("data/fenz_depths.rds") |> 
  dplyr::filter(!is.na(z_max))
n_fenz <- nrow(fenz_depths)

library(ggplot2)

```

## Morphometric distribution of depth and basin shape in Aotearoa New Zealand lakes

Tadhg Moore

## Overview

-   Background
-   Methods
-   Results
-   Conclusions
-   Future Work

## Motivation

-   Developing the Lake Ecosystem Research New Zealand modelling platfomr (LERNZmp).
-   Lake depth and morphometry are key drivers of lake ecosystem function, therefore accurate depth data is essential.

## Background {.smaller}

::::: columns
::: {.column width="50%"}
-   Lack of a central database on lake morphometry, particularly depth
-   Large number of bathymetric data has been recorded across lakes in New Zealand
-   Current predictions of max lake depth, within the FENZ database (n=`r n_fenz`) are limited and have clear biases
:::

::: {.column width="50%"}
```{r}
#| label: fig-fenz-depth-distribution
#| fig.cap: "Depth distribution of FENZ lakes"
#| fig.width: 6
#| fig.height: 6

ggplot() +
  geom_histogram(data = fenz_depths, aes(x = z_max),
                 bins = 50, colour = "white") +
  # geom_point(data = fenz_depths, aes(x = area_Ha, y = z_max)) +
  scale_x_log10(breaks = c(1, 5, 10, 20, 50, 100, 400)) +
  # scale_y_log10() +
  labs(y = "Count", x = "Depth (m)")+
  theme_bw(base_size = 16) 

# library(plotly)
# plot_ly() |> 
#   add_histogram(data = fenz_depths, x = ~z_max, nbins = 50) |>
#   layout(xaxis = list(title = "Depth (m)", type = "log"),
#          yaxis = list(title = "Frequency"))
# add_markers(data = fenz_depths, x = ~area_Ha, y = ~z_max,
#             opacity = 0.3) |> 
# layout(xaxis = list(type = "log", title = "Area (ha)"),
#        yaxis = list(type = "log", title = "Depth (m)"))

```
:::
:::::

## Max depth vs Lake area

```{r}
#| label: fig-fenz-depth-area
#| fig.cap: "Depth vs Area of FENZ lakes"

ggplot() +
  geom_point(data = fenz_depths, aes(x = area_Ha, y = z_max),
             alpha = 0.2, size = 1) +
  scale_y_log10(breaks = c(1, 5, 10, 20, 50, 100, 400)) +
  scale_x_log10(breaks = c(1, 5, 10, 50, 100, 1000)) +
  labs(x = "Area (ha)", y = "Depth (m)")+
  theme_bw(base_size = 16) 

```

<!-- ## Locations of lakes with bathymetry data in New Zealand {.smaller} -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| include: false -->
<!-- #| label: get-nz-regions-shapefile -->

<!-- # Read the shapefile data from Stats NZ -->
<!-- url <- "https://datafinder.stats.govt.nz/" # Base URL -->
<!-- layer_id <- 111182 # Layer ID for the regional shapefile -->

<!-- # Read the shapefile data -->
<!-- nz_shapefile <- aemetools::read_web_sf(url = url, layer_id = layer_id, -->
<!--                                        key = Sys.getenv("STATS_NZ_KEY")) -->

<!-- # Remove the Chatham Islands from the shapefile -->
<!-- nz_shapefile <- nz_shapefile |> -->
<!--   dplyr::filter(C2023_V != "Area Outside Region") -->

<!-- # Read the Coast shapefile data from LINZ -->
<!-- url <- "https://data.linz.govt.nz/" # Base URL -->
<!-- layer_id <- 51153 # Layer ID for the national coast line -->

<!-- # Read the shapefile data -->
<!-- nz_coast <- aemetools::read_web_sf(url = url, layer_id = layer_id) -->
<!-- nz_coast <- nz_coast |>  -->
<!--   sf::st_transform(2193) -->

<!-- nz_coast <- sf::st_crop(nz_coast, nz_shapefile) -->
<!-- nz_coast_simp <- sf::st_simplify(nz_coast, dTolerance = 100) -->

<!-- nz_region_clip <- sf::st_intersection(nz_shapefile, nz_coast_simp) |>  -->
<!--   dplyr::mutate(legend_name = gsub(" Region", "", C2023_V), -->
<!--                 legend_name = gsub("\\?", "ū", legend_name)) -->

<!-- # tm_shape(sub) + -->
<!-- #   tm_borders(col = "black", lwd = 2) -->

<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: fig-fenz-map -->
<!-- #| fig.cap: "Map of FENZ lakes" -->

<!-- library(tmap) -->
<!-- tmap_mode("view") -->
<!-- tm_shape(nz_region_clip) + -->
<!--   tm_borders(col = "black", lwd = 2)  -->

<!-- # tm_shape(fenz_depths) + -->
<!-- #   tm_polygons(col = "z_max", style = "cont") + -->
<!-- #   tm_layout(legend.position = c("right", "bottom")) + -->
<!-- #   tm_view() -->
<!-- ``` -->

## Methods {.smaller}

-   Bathymetry survey data were sourced from a variety of sources - paper maps, digital files, and online databases (n=.
-   Shoreline data were updated using satellite and aerial imagery.
-   Depth data were interpolated using a Multilevel B-Spline Approximation (MBA) algorithm.
-   Hypsographic curves were calculated for each lake.
-   Comparison of depth and area with other databases.
-   Used machine learning to predict lake depth from morphometric data for all lakes in New Zealand and benchmark against the FENZ dataset.

## Results {.r-fit-text}

```{r}
#| label: model-results

max_depth_r2 <- 0.8
mean_depth_r2 <- 0.6

fenz_max_depth_r2 <- 0.4
fenz_mean_depth_r2 <- 0.2

```

Bathymetric data from 156 lakes were digitised and collated for use in this study.

Using a machine learning model, we were able to predict maximum lake depth from morphometric data with an $r^2$ = `r max_depth_r2` and mean lake depth with an $r^2$ = `r mean_depth_r2`.

This was substantially better than the FENZ dataset, which had an $r^2$ = `r fenz_max_depth_r2` for max depth and `r fenz_mean_depth_r2` for mean depth.

## Predicting mean and max depth {.smaller}

:::::: panel-tabset
### Max depth

::::: columns
::: {.column width="50%"}
Figure
:::

::: {.column width="50%"}
Table of fit
:::
:::::

### Mean depth

Content for `Tab B`
::::::
